MolProbity's Ultimate Rotamer-Library Distributions for Model Validation

### 2016-02-25 introduction.tex
Small changes befoor resubmition

Because rotamer libraries or distributions are an integral component of modern structural biology, it is imperative that they provide only authentic, low-energy sidechain conformations so that errors are not propagated. Such errors of circular reasoning can be documented \cite{lovell2000penultimate} for many early cases, such as high-energy eclipsed $\chi$ angles added for "completeness", or real empirical data clusters caused by incorrect backward-fit branched sidechains. The accuracy of these libraries depends on including only reliably modeled sidechains in the reference dataset. All rotamer libraries filter their datasets at the file level by resolution and redundancy. As the PDB grew in size, it  became increasingly practical to filter also at the residue level. This process very effectively lowers noise and sharpens clustering, because even at high resolution the poorly ordered regions are susceptible to misfitting and are often worse than the good parts at low resolution.

>>>>

Because rotamer libraries or distributions are an integral component of modern structural biology, it is imperative that they provide only authentic, low-energy sidechain conformations so that errors are not propagated. Such errors of circular reasoning can be documented for many early cases \cite{lovell2000penultimate}, such as high-energy eclipsed $\chi$ angles added for "completeness", or real empirical data clusters caused by incorrect backward-fit branched sidechains. The accuracy of these libraries depends on including only reliably modeled sidechains in the reference dataset. All rotamer libraries filter their datasets at the file level by resolution and redundancy. As the PDB grew in size, it  became increasingly practical to filter also at the residue level. This process very effectively lowers noise and sharpens clustering, because even at high resolution the poorly ordered regions are susceptible to misfitting and are often worse than the good parts at low resolution.

### 2016-02-25 methods.tex
Small changes befoor resubmition

All rotamer statistical analysis presented here used the Top8000-SFbest\_hom70 dataset version. Deposited structure factors not only are required for electron-density based residue filtering, but they also enable examination of the model in the density map for individual examples of conformations deemed dubious or interesting. Of the 7,419 chains, 28 failed in the RSCC calculation described below, and 175 were dropped because $\le$ 20\% of their residues remained after residue-level filtering. The final rotamer dataset of 7,216 chains is listed on GitHub at \texttt{Top8000/Top8000\_rotamer\_pdb\_chain\_count.csv}.

@@

In selecting filter thresholds for the dataset, we had the goal of keeping a large number of residues while reliably eliminating those with dubious density. To do this, we analyzed the counts of residues remaining for several combinations of filter thresholds. The best balance between filter levels and a reasonably large number of residues yielded the following thresholds: \textcolor{changecolor}{worst per-atom isotropic} B factor $<$ 40 \AA\textsuperscript{2}, worst correlation coefficient $>$ 0.7, and worst map value $>$ 1.1 $\sigma$. Upon visual inspection of residues and maps close to these filter thresholds, we determined that the selected combination of thresholds were indeed effective at keeping only residues with satisfactory electron density. Additionally, residues were required to have no all-atom clashes, an occupancy of 1.0, B factors $>$ 1.0 \AA\textsuperscript{2} and all backbone atoms modeled, achieved indirectly by ensuring that $\phi$, $\psi$, $\omega$ and $\tau$ were defined for each residue.  The latter test also drops sidechains of the first and last residue in each chain, which are somewhat less and differently affected by backbone interactions.

@@

\subsection{Rotamer Assignment and Score}
\label{sec_md:rotaassign}
For the purpose of setting the appropriate contour level for the new outlier cutoff, a large test dataset was created from all PDBs in the Top8000 by including all chains and all residues in each PDB file (as a surrogate for entire new PDB files later being validated), hereafter called the unfiltered dataset. \textit{phenix.rotalyze} was used to assign rotamer names and scores for each residue in the unfiltered dataset. The score is calculated from the individual residue's $\chi$ values, by interpolating over the nearest contour grid points (see section \ref{sec_md:contours}). This was done twice, once using the new Top8000 contours and once using our previous Top500 contours. If the score is above the outlier threshold, then a rotamer name is assigned based on which rotamer well the given $\chi$ angles fall within (the maximum well ranges were described in Section \ref{sec_md:chi_stats}). Otherwise the side-chain conformation is classified as an \textit{outlier.}  \textcolor{changecolor}{A Top8000 cutoff of 0.3\% flagged very nearly the same number of outliers in the unfiltered dataset as the current Top500 cutoff of 1.0\% , so the value of 0.3\% was chosen as the outlier contour level for the new system.}

>>>>

All rotamer statistical analysis presented here used the Top8000-SFbest\_hom70 dataset version. Deposited structure factors not only are required for electron-density based residue filtering, but they also enable manual examination of the model in the density map for individual examples of conformations deemed dubious or interesting. Of the 7,419 chains, 28 failed in the RSCC calculation described below, and 175 were dropped because $\le$ 20\% of their residues remained after residue-level filtering. The final rotamer dataset of 7,216 chains is listed on GitHub at \texttt{Top8000/Top8000\_rotamer\_pdb\_chain\_count.csv}.

@@

In selecting filter thresholds for the dataset, we had the goal of keeping a large number of residues while reliably eliminating those with dubious density. To do this, we analyzed the counts of residues remaining for several combinations of filter thresholds. The best balance between filter levels and a reasonably large number of residues yielded the following thresholds: \textcolor{changecolor}{worst per-atom isotropic} B factor $<$ 40 \AA\textsuperscript{2}, worst correlation coefficient $>$ 0.7, and worst map value $>$ 1.1 $\sigma$. Upon visual inspection of residues and maps close to these filter thresholds, we determined that the selected combination of thresholds were indeed effective at keeping only residues with satisfactory electron density. Additionally, residues were required to have no all-atom clashes, an occupancy of 1.0, B factors $>$ 1.0 \AA\textsuperscript{2} and all backbone atoms modeled, achieved indirectly by ensuring that $\phi$, $\psi$, $\omega$ and $\tau$ were defined for each residue.  The latter test also drops first and last residue in each PDB chain, which are somewhat less and differently affected by backbone interactions.

@@

\subsection{Rotamer Assignment and Score}
\label{sec_md:rotaassign}
For the purpose of setting the appropriate contour level for the new outlier cutoff, a large test dataset was created from all PDBs in the Top8000 by including all chains and all residues in each PDB file (as a surrogate for entire new PDB files later being validated), hereafter called the unfiltered dataset. \textit{phenix.rotalyze} was used to assign rotamer names and scores for each residue in the unfiltered dataset. The score is calculated from the individual residue's $\chi$ values, by interpolating over the nearest contour grid points (see section \ref{sec_md:contours}). This was done twice, once using the new Top8000 contours and once using our previous Top500 contours. \textcolor{changecolor}{Outlier counts were made at several Top8000 contour cutoffs and then compared to the outlier counts as determined by the Top500. In order to keep outlier counts approximately equal between the previous and new systems, a Top8000 outlier cutoff of 0.3\% was selected.} If the score is above the outlier threshold, then a rotamer name is assigned based on which rotamer well the given $\chi$ angles fall within (well ranges were described in Section \ref{sec_md:chi_stats}). Otherwise the side-chain conformation is classified as an \textit{outlier.}

### 2016-02-25 results.tex
Small changes befoor resubmition

\section{Results}
In the Top8000 system there are a total of \textcolor{changecolor}{212 named rotamer clusters, compared with 153 in our 2000 paper [1]}. Many of the central values have shifted somewhat from the Top500 system, both because of much more and higher-quality data and also because of the new center-of-mass definition (see Section \ref{sec_md:chi_stats}), which we feel represents the clusters better than either the modal or the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: Table S3 describes the frequency and count of each rotamer in each residue type and Tables S4-S21 give central dihedral and covalent angle values for each rotamer. These numbers are available in CSV format on GitHub under \texttt{Top8000/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/reference_data}. The more useful multi-dimensional distributions are also on GitHub, as linked in Section \ref{sec_md:contours}.

>>>>

\section{Results}
In the Top8000 system there are a total of \textcolor{changecolor}{212 named rotamer clusters, compared with 153 in our 2000 paper [1]}. Many of the central values have shifted somewhat from the Top500 system, both because of much more, higher-quality data and because of the new center-of-mass definition (see Section \ref{sec_md:chi_stats}), which we feel represents the clusters better than either the modal or the common-atom values defined previously \cite{lovell2000penultimate}. MolProbity's "ultimate" rotamers and their parameters are given in two tables: Table S3 describes the frequency and count of each rotamer in each residue type and Tables S4-S21 give central dihedral and covalent angle values for each rotamer. These numbers are available in CSV format on GitHub under \texttt{Top8000/Top8000\_rotamer\_central\_values} at \url{http://github.com/rlabduke/reference_data}. The more useful multi-dimensional distributions are also on GitHub, as linked in Section \ref{sec_md:contours}.
